Climate-driven phytoplankton community variability in the Cariaco Basin, Venezuela

### 2025-05-27 agujournaltemplate.tex
updated Methods and Results almost done

    Data sets for Niskin bottle samples, CTD, and phytoplankton taxonomy were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047} (accessed in February, 2024). 
    From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps (i.e. more than three months) in coverage. Therefore, we did not include zooplankton net tow and HPLC pigment data in our analysis. Our study instead focused on the phytoplankton microscopy data and bottom-up drivers of change. We compared these data with Niskin bottle nutrient, temperature, and salinity data, as well as climate data and indices, focusing on bottom-up drivers of community variability and diversity. In total, we used a complete set of 254 monthly data points, on which all variables were available.

\subsection{Climate data}
    %from Taylor et al. 2012
    Zonal wind stress (-u, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters via Ekman transport \cite{rueda-roa_southern_2013}. We extracted the -u wind component at 10 m from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis in the grid cell closest to the Cariaco time series sampling coordinates (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels} (accessed June 2024). This variable is referred to as "wind speed" throughout the manuscript for simplicity.

@@

    \subsubsection{Data processing}
    % Niskin, CTD and Phytoplankton data processing
    Data were processed and analyzed using the R programming language (version 4.3.3) \cite{r_core_team_r_2024} with data pipelines created using tidyverse packages \cite{wickham_welcome_2019}. All data processing and analysis code is publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.
    For all discrete depth measurements, the data were interpolated using the "oce" package \cite{kelley_oce_2023} with the adaptive "unesco" algorithm defined by the U.S. National Oceanographic Data Center \cite{johnson2006world}. The interpolated values were depth integrated to 100 meters and in four discrete intervals between 0-25, 25-50, 50-75 and 75-100 \meter for further analysis. Additionally, sea surface temperature (SST) was calculated from CTD temperature measurements averaged over the top \qty{10}{\meter}. For phytoplankton microscopy data, individually identified taxonomic units (species or genus) were interpolated, integrated, and then grouped by genus via the sum of interpolated counts.

@@

    \subsubsection{Phytoplankton Community Structure Analysis}
    To investigate the structure of the phytoplankton time series data, we used a complete linkage clustering method via the "hclust" function in R \cite{r_core_team_r_2024} on the binary Jaccard distance matrix of annual aggregates of observed genera integrated over 100 m depth. Additionally, we performed non-metric multidimensional scaling (NMDS) ordination and hierarchical clustering on the monthly integrated counts that were square-root transformed using "metaMDS" from the "vegan" package \cite{oksanen_vegan_2024}. We employed the Bray distance matrix in a two-dimensional ordination that reached a converging solution within 30 steps. To relate the multidimensional structuring of the community data to variability in environmental data, we used the "envfit" function from the "vegan" package to fit the environmental vectors onto the ordination with 999 permutations. The arrows indicate the gradient of a two-dimensional linear trend surface. 
    To contrast the environmental parameters and diversity data between the community clusters resulting from the yearly aggregated clustering, we created density distribution plots contrasting the monthly measurements between clusters using the "ggpubr" package \cite{kassambara_ggpubr_2023}. To test whether the environmental variables, chlorophyll $a$ and diversity metrics varied significantly between the two community clusters, we employed a two-sided Wilcoxon sum rank test using the "wilcox.test" function in R.
    % detailed explanation of NMDS in Winder and Hunter 2008

    \subsubsection{Assessing Environmental Drivers of Phytoplankton Dynamics}
    To quantify the effects of bottom-up drivers in the Cariaco Basin on shifts in the phytoplankton community, we employed a Gradient Forest analysis. Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables (here: monthly observed genus counts) to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}. Gradient Forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then averages these to determine the overall importance of each predictor. The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for phytoplankton community turnover. Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}. This analysis was conducted using monthly measurements integrated to 100 m depths for observed genus counts, ensuring the inclusion of all available data. The drivers/predictor variables considered in the analysis were the AMO index, the MEI v.2 index, the -u 10m wind component, the \qty{21}{\celsius} isotherm, SST, and salinity, as well as the concentration of nutrients ($NO_3$, $PO_4$ and $SiO_4$). The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology takes into account by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}. To consider the temporal lag in effect, particularly of climate indices, we included time lags of up to 6 months for all climate variables, as well as 3-month lags for in-situ measurements in exploratory runs. We chose time lags that scored highest in the goodness-of-fit statistic for the final model run (see Table \ref{app:table1} in the Appendix).
    % Note: Add some more info about Supplmentary Results and test runs here!

@@

\caption{All data sources with corresponding units used as predictor variables for the gradient forest analysis.}
\centering
\resizebox{\textwidth}{!}{\begin{tabular}[t]{rrl}
\toprule
Variable Name & Units  & Data source \\
\midrule
MEI v.2 & unitless & Climate index provided by NOAA \\
AMO & unitless & Climate index provided by NSF NCAR \\
Wind speed & m/s & ERA5 model reanalysis, negative 10 m u-component of wind speed \\
Precipitation & m & ERA5 model reanalysis, total precipitation\\
Evaporation & m & ERA5 model reanalysis, evaporation \\
SST & °C & CTD, Average temperature measured in top 10 m\\
21° C Isotherm & m & CTD, depth at which temperature reaches 21°C\\
Salinity & PSU & Niskin, top 100m \\
NO3 & µM & Niskin, top 100m \\
PO4 & µM & Niskin, top 100m \\
SiO4 & µM & Niskin, top 100m \\
\end{tabular}}
\label{tab:DataSources}
\end{table}

@@

% Figure 2
The normalized yearly means for all environmental variables considered, as well as the logarithmic transformation of chlorophyll \textit{a} and phytoplankton cell counts per functional group, are shown in Figure \ref{fig:zscore}. We included marks to highlight those years with a significant number of missing monthly measurements (\textgreater 2 months), as these make the annual means less reliable. Particularly towards the later years of the time series, the sampling became sparse. The patterns emerging from wind speed, SST, and the directly related \qty{21}{\celsius} isotherm depth clearly show the years of positive anomalies of wind (blue) and negative anomalies for SST (red) and upwelling (red, i.e., shallower depth of the isotherm). These patterns are reflected in the mode of the AMO, where a positive anomaly seems to correlate with weak upwelling regimes (see figure \ref{fig:sup:correlation} in the appendix). The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series but show differing inter-annual oscillations. Annual mean anomalies of nitrate and phosphate seem to follow upwelling anomalies but remain relatively low between 2002 and 2014, with some variability in mean phosphate concentration. Silicate does not follow a clear pattern but shows an elevated mean concentration toward the later part of the time series (between 2008 and 2013), which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion via diminished diatom abundances.

% Figure 3
To further investigate the depth distribution and dynamics of chlorophyll \textit{a} and community diversity, Figure \ref{fig:divts} shows the depth distributions over time, as well as between the rainy and upwelling seasons. The bulk of the chlorophyll \textit{a} biomass ($\sim$ 80\%) was measured in the upper 50 meters of the water column, with approximately 15\% measured between 50 and 75 meters. Chlorophyll \textit{a} shows marked seasonality, with the highest biomass in the upper water column during the winter months (December–February, see Figure \ref{fig:divts} $b$). The yearly mean chlorophyll \textit{a} biomass shows the highest variability within the upper 25 meters. In some years, particularly during 2000, 2005, 2007, 2008, and 2013, its concentration in this layer was lower than in the 25–50 meter depth range (see Figure \ref{fig:divts} $a$). In general, yearly mean anomalies within the upper 75 meters appear to be elevated during the period 1996 to 2004 and only recover in 2012. The year 2013 is an outlier, particularly for the upper 25 meters, as the missing monthly samplings from the winter months (January and February) skew the value. Nevertheless, the monthly measurements for the remainder of the year 2013 show a markedly lower biomass (65\% of the mean over the entire time series for these months), which recovers in 2014. Chlorophyll \textit{a} biomass below 50 meters depth shows a relatively stable distribution throughout the year and across the time series. In fall, during the rainy season, mean chlorophyll \textit{a} biomass in the upper 25 meters is markedly reduced, while the 50–75 meter depth interval sees a slight increase, indicating a shift in the chlorophyll \textit{a} maximum to lower depths under weak upwelling conditions.
The calculated genus diversity from the microscopy cell counts generally follows the dynamics of chlorophyll \textit{a} biomass in seasonality, with maximum values observed during the upwelling season in winter. Genus richness is consistently highest in the top \qty{25}{meters} depth interval, which also shows the most variability over the time series. Across the time series, we observe maximum richness at the start of the time series (1996-1997), with a subsequent reduction in diversity across all depths, which only recovers after 2011. The collapse in diversity following the start of the time series does not follow the pattern shown by chlorophyll \textit{a} biomass, which remains relatively stable across the top \qty{100}{meters} until 2004 (see Figure \ref{fig:divts $a$}). Scaling this genus diversity by the cell count abundances, i.e. the Shannon index, shows a similar but less pronounced dynamic. Diversity remains high throughout 2000 and recovers more quickly, but shows marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009, and 2016. The seasonal signal is almost lost (see Figure \ref{fig:divts} $f$). 
Community evenness as signified by the Pielou index shows no seasonal signal and a very large variance within seasons (see Figure \ref{fig:divts} $h$). The inter-annual mean dynamics of evenness are remarkably stable and similar across all depth intervals, clustered around an intermediate 0.5 value. There are marked reductions in all depths towards 2000 and for all depths above 75 meters in 2006, 2009 and 2016. Throughout 2003 to 2013 evenness is almost consistently highest in the deepest interval below 75 meters, where the community seems to have been more evenly distributed.

% Figure 4
Taking the microscopy community data, interpolated within the top 100 meters, we wanted to find if there was an apparent structure within the time series that could be related to the physical environment. To achieve this, we employed clustering analysis, both on the annual aggregate of cell counts from microscopy and on the individual monthly measurements. Figure \ref{fig:clustering} $a$ shows the results of clustering the annual aggregate communities with a linear complete linkage algorithm based on the binary Jaccard matrix. The community is divided into two groups, with the first cluster including the years 1996-2004 and 2014-2016 and the second cluster of the intermediate years from 2004-2013. 
Inter-annual species turnover is shown in Figure \ref{fig:clustering} $b$, where there is an increasing trend in turnover until 2004, where turnover is reduced for the period of cluster 2, but in 2014 with the return to cluster 1, the turnover is higher again. This pattern is also reflected in the clustering in Figure \ref{fig:clustering} $a$, where cluster 2 shows a lower distance in the clustering tree between the years.

To check whether annual aggregation of the community data smoothed out strong seasonal signals, we performed an NMDS cluster analysis using the monthly cell counts (see Figure \ref{fig:clustering} $c$). This NMDS clustering is overlaid with environmental data and climate indices to help explain some of the underlying trends. The first axis separates the two clusters, where cluster 2 is trending towards positive AMO anomalies and cluster 2 trends towards positive anomalies in salinity and the MEI v.2 index. The second NMDS axis appears to follow the trends in upwelling intensity as described by the opposing trend arrows for nutrient concentrations and temperature, as well as wind speed. Overall, NMDS clustering seems to confirm the pattern found by the annual aggregate clustering. 
Figure \ref{fig:clustering} $d$ to $o$ shows density distributions for environmental variables, climate and diversity indices between the two clusters. Upwelling related variables show distinct patterns between the two clusters. For example, cluster 1 experiences more frequent high wind events, a shallower \qty{21}{\celsius} Isotherm, lower temperature, and higher salinity and nutrient concentrations, in particular for nitrate. The AMO index shows a separation between the two clusters, where cluster 1 shows more negative anomalies, while cluster 2 has a high occurrence of positive anomalies. Although the mean diversity for all the indices is relatively similar, with a slightly higher mean in the Shannon and Pielou index for cluster 2, the density distribution shows marked differences. For cluster 1 there is a higher variance in genus richness, while cluster 2 shows higher variance and a bi-modal distribution for Shannon and Pielou indices.

% Figure 5
To identify the variables that best predict the shifts observed within the phytoplankton community, we performed a Gradient Forest analysis on the microscopy cell counts interpolated over the top 100 meters and grouped by identified genus. Preliminary model runs were used to identify if time lags of the predictor variables could better explain the data (see Appendix), and this resulted in choosing the AMO with a lag of 2 months and MEI v.2 with a lag of 4 months. In Figure \ref{fig:GF} $a$, we show the predictor variables ranked by their weighted importance, with the highest predictive capacity for climate indices with time lags and upwelling-related variables such as nitrate in the upper 100 meters and wind speed scoring slightly lower. The \qty{21}{\celsius} Isotherm and silicate and phosphate concentrations scored lowest in predicting changes within the community. The way Gradient Forest predicts shifts in the community is by building a random forest model for each observed genus across the time series. The predictive power of the models for each genus is shown in Figure \ref{fig:GF} $b$. The microscopy data are biased towards larger cell sizes, and therefore most of the predicted genus cell counts are from diatoms. However, two Haptophytes and four Dinoflagellates, as well as the Cyanobacteria \textit{Synechococcus} do rank highly.
Figure \ref{fig:GF} $c$ shows the cumulative importance of the predictor variables across the parameter range. The black line indicates cumulative importance across the entire community and the colored dashed lines indicate importance per phytoplankton group. For the AMO index, we see a clear pattern in which negative anomalies highly affect the community. This contrasts with a higher importance for positive anomalies in the MEI v.2 index, particularly for haptophytes. Nitrate concentrations between 6 and 8 \unit{\micro \mole} is where the community changed most, indicating these values as a threshold for the composition of the community. For wind speed, temperature, and salinity, changes occurred most at parameter ranges related to high upwelling, i.e., high wind (between -4 and -6 \unit{m.s^{-1}}), low temperature (below \qty{23}{\celsius}) and high salinity (above \qty{36.8}{PSU}). An interesting aspect is that, for both climate indices, haptophytes showed the strongest cumulative importance, but for diatoms, the nitrate concentration seemed to be a stronger predictor.

% ADD HERE MORE DISCUSSION OF MODEL COMPARISON

@@

\begin{figure}
\begin{center}
\noindent\includegraphics[width=480pt]{fig/DepthDivTimeSeries_updated.pdf}
\end{center}
\caption{Time series and seasonality of chlorophyll \textit{a} and diversity indices across four depth intervals. (a,c,e,g) Annual mean across time series for years with more than 5 monthly sampling. Transparent dots show monthly raw values. (b,d,f,h) Boxplots of monthly measurements across four depth intervals grouped by season. Upwelling season lasts from December to April, Rainy Season from May to November.}
\label{fig:divts}
\end{figure}

@@

This study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which is, in turn, influenced by large-scale climatic changes. In the Cariaco Basin, decadal-scale climate oscillations have impacted the wind-driven upwelling regime, resulting in confounding effects on the composition and biodiversity of the phytoplankton community.
% - a few sentences on basic seasonal dynamics removed here
These changes are well reflected in the yearly mean anomalies (see Figure \ref{fig:zscore}) and the density distribution of monthly measurements (see Figure \ref{fig:clustering}). After 2004, we observed higher sea surface temperatures (SST), lower wind speeds, reduced fluorometric chlorophyll \textit{a}, and a collapse in functional group cell counts, all corresponding with a shift in the Atlantic Multidecadal Oscillation (AMO) toward positive mean anomalies until 2014. 
% - a few sentences about shift to smaller taxa and collapse in Sardine Fisheries

>>>>

    Data sets for Niskin bottle samples, CTD, and phytoplankton taxonomy were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047} (accessed in February, 2024). 
    From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps (i.e. no more than three months) in coverage. Therefore, we did not include zooplankton net tow and HPLC pigment data in our analysis. Our study instead focused on the phytoplankton microscopy data and bottom-up drivers of change. We compared these data with Niskin bottle nutrient, temperature, and salinity data, as well as climate data and indices, focusing on bottom-up drivers of community variability and diversity. In total, we used a complete set of 254 monthly data points, on which all variables were available.

\subsection{Climate data}
    %from Taylor et al. 2012
    Zonal wind stress (-u, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters via Ekman transport \cite{rueda-roa_southern_2013}. We extracted the -u wind component at 10 m from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis in the grid cell closest to the Cariaco time series sampling coordinates (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels} (accessed June 2024). This variable is referred to as "wind speed" throughout the manuscript for simplicity. The data for precipitation and evaporation are similarly extracted from the ERA5 model reanalysis dataset for the aforementioned grid cell. Evaporation provides an estimate of the accumulated amount of water that has evaporated from the Earth's surface, while precipitation is the estimate of total precipitation, comprising all liquid that falls to Earth's surface as rain or snow.

@@

    \subsubsection{Data processing}
    % Niskin, CTD and Phytoplankton data processing
    Data were processed and analyzed using the R programming language (version 4.3.3) \cite{r_core_team_r_2024} with data pipelines created using tidyverse packages \cite{wickham_welcome_2019}. All data processing and analysis code is publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.
    For all discrete depth measurements, the data were interpolated using the "oce" package \cite{kelley_oce_2023} with the adaptive "unesco" algorithm defined by the U.S. National Oceanographic Data Center \cite{johnson2006world}. The interpolated values were depth integrated to 100 meters and in four discrete intervals between 0-25, 25-50, 50-75 and 75-100 \meter for further analysis. Additionally, sea surface temperature (SST) was calculated from CTD temperature measurements averaged over the top \qty{10}{\meter}. For phytoplankton microscopy data, individually identified taxonomic units (species or genus) were interpolated and integrated across depth, and finally grouped by genus via the sum of interpolated counts.

@@

    \subsubsection{Phytoplankton Community Structure Analysis}
    To investigate the structure of the phytoplankton time series data, we used a complete linkage clustering method via the "hclust" function in R \cite{r_core_team_r_2024} on the binary Jaccard distance matrix of annual aggregates of observed genera integrated over 100 m depth. Additionally, we performed non-metric multidimensional scaling (NMDS) ordination and hierarchical clustering on the monthly integrated counts that were square-root transformed using "metaMDS" from the "vegan" package \cite{oksanen_vegan_2024}. We employed the Bray distance matrix in a two-dimensional ordination that reached a converging solution within 30 steps. To relate the multidimensional structuring of the community data to variability in environmental data, we used the "envfit" function from the "vegan" package to fit the environmental vectors onto the ordination with 999 permutations. The arrows indicate the gradient of a two-dimensional linear trend surface. 
    To contrast the environmental parameters and diversity data between the community clusters resulting from the yearly aggregated clustering, we created density distribution plots contrasting the monthly measurements between clusters using the "ggpubr" package \cite{kassambara_ggpubr_2023}. To test whether the environmental variables, chlorophyll $a$ and diversity metrics varied significantly between the two community clusters, we employed a two-sided Wilcoxon sum rank test using the "wilcox.test" function in R \cite{r_core_team_r_2024}.
    % detailed explanation of NMDS in Winder and Hunter 2008

    \subsubsection{Assessing Environmental Drivers of Phytoplankton Dynamics}
    To quantify the effects of bottom-up drivers in the Cariaco Basin on shifts in the phytoplankton community, we employed a Gradient Forest analysis. Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables (here: monthly observed genus counts) to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}. Gradient Forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then averages these to determine the overall importance of each predictor. The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for phytoplankton community turnover. Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}. This analysis was conducted using monthly measurements integrated to 100 m depths for observed genus counts, ensuring the inclusion of all available data. The drivers/predictor variables considered in the analysis were the AMO index, the MEI v.2 index, wind speed, evaporation, precipitation, the \qty{21}{\celsius} isotherm, SST, and salinity, as well as the concentration of nutrients ($\mathrm{NO}_3$, $\mathrm{PO}_4$ and $\mathrm{SiO}_4$). 
    The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology accounts for by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}. 
    We show a model run without any time lags in the supplemental material (see Figure \ref{fig:sup:GFoutput_nolags}). To consider the temporal lag in effect, particularly of climate indices, we included time lags of up to 6 months for all climate variables, as well as 3-month lags for in-situ measurements in exploratory runs. We chose time lags that scored highest in the goodness-of-fit statistic for the final model run (see Table \ref{app:table1} in the Appendix). 
    % Note: Perhaps add some more info about Supplmentary Results and test runs here?

@@

\caption{All data sources with corresponding units used as predictor variables for the gradient forest analysis.}
\centering
\resizebox{\textwidth}{!}{\begin{tabular}[t]{rrl}
\toprule
Variable Name & Units  & Data source \\
\midrule
MEI v.2 & unitless & Climate index provided by NOAA \\
AMO & unitless & Climate index provided by NSF NCAR \\
Wind speed & \qty{}{\meter/\second} & ERA5 model reanalysis, negative 10 m u-component of wind speed \\
Precipitation & \qty{}{\meter} & ERA5 model reanalysis, total precipitation\\
Evaporation & \qty{}{\meter} & ERA5 model reanalysis, evaporation \\
SST & \qty{}{\celsius} & CTD, Average temperature measured down to \qty{10}{\meter} depth\\
\qty{21}{\celsius} isotherm & \qty{}{\meter} & CTD, depth at which temperature reaches 21°C\\
Salinity & PSU & Niskin, interpolated down to \qty{100}{\meter} depth \\
$\mathrm{NO}_3$  & µM & Niskin, interpolated down to \qty{100}{\meter} depth \\
$\mathrm{PO}_4$  & µM & Niskin, interpolated down to \qty{100}{\meter} depth \\
$\mathrm{SiO}_4$  & µM & Niskin, interpolated down to \qty{100}{\meter} depth \\
\end{tabular}}
\label{tab:DataSources}
\end{table}

@@

% Figure 2
The normalized yearly means for all environmental variables considered, as well as the logarithmic transformation of chlorophyll \textit{a} and phytoplankton cell counts per functional group, are shown in Figure \ref{fig:zscore}. We included marks to highlight those years with a significant number of missing monthly measurements (\textgreater 2 months and $\geq$ 4 months), as these make the annual means less reliable. Particularly toward the later years of the time series, the sampling became sparse. The patterns emerging from wind speed, evaporation, precipitation, SST, and \qty{21}{\celsius} isotherm depth clearly show the years of positive anomalies of wind (blue) and negative anomalies for SST (red) and upwelling (red, i.e., shallower depth of the isotherm). These patterns are reflected in the mode of the AMO, where a positive anomaly seems to correlate with weak upwelling regimes (see Figure \ref{fig:sup:correlation} in the appendix). The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series but show differing inter-annual oscillations. Annual mean anomalies of nitrate and phosphate seem to follow upwelling anomalies but remain relatively low between 2002 and 2014, with some variability in mean phosphate concentration. Silicate does not follow a clear pattern but shows an elevated mean concentration toward the later part of the time series (between 2008 and 2013), which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion due to diminished diatom abundances.

% Figure 3
To further investigate the depth distribution and dynamics of chlorophyll \textit{a} and community diversity, Figure \ref{fig:divts} shows the depth distributions over time, as well as between the rainy and upwelling seasons. The bulk of the chlorophyll \textit{a} biomass ($\sim$ 80\%) was measured in the upper 50 meters of the water column, with approximately 15\% measured between 50 and 75 meters. Chlorophyll \textit{a} exhibits marked seasonality, with the highest biomass in the upper water column during the winter months (December–February, see Figure \ref{fig:divts} $b$). The yearly mean chlorophyll \textit{a} biomass shows the highest variability within the upper 25 meters. In some years, particularly during 2000, 2005, 2007, 2008, and 2013, its concentration in this layer was lower than in the 25–50 meter depth range (see Figure \ref{fig:divts} $a$). In general, yearly mean anomalies within the upper 75 meters appear to be elevated during the period from 1996 to 2004 and only recover in 2012. The year 2013 is an outlier, particularly for the upper 25 meters, as the missing monthly samplings from the winter months (January and February) skew the value. Nevertheless, the monthly measurements for the remainder of the year 2013 show a markedly lower biomass (65\% of the mean over the entire time series for these months), which recovers in 2014. Chlorophyll \textit{a} biomass below 50 meters depth shows a relatively stable distribution throughout the year and across the time series. In fall, during the rainy season, mean chlorophyll \textit{a} biomass in the upper 25 meters is markedly reduced, while the 50–75 meter depth interval sees a slight increase, indicating a shift in the chlorophyll \textit{a} maximum to lower depths under weak upwelling conditions.
The calculated genus diversity from the microscopy cell counts generally follows the dynamics of chlorophyll \textit{a} biomass in seasonality, with maximum values observed during the upwelling season in winter. Genus richness is consistently highest in the top \qty{25}{meters} depth interval, which also shows the most variability over the time series. Across the time series, we observe maximum richness at the start of the time series (1996-1997), with a subsequent reduction in diversity across all depths, which only recovers after 2011. The collapse in diversity following the start of the time series does not follow the pattern shown by chlorophyll \textit{a} biomass, which remains relatively stable across the top \qty{100}{meters} until 2004 (see Figure \ref{fig:divts $a$}). Scaling this genus diversity by the cell count abundances, i.e. the Shannon index, shows a similar but less pronounced dynamic. Diversity remains high throughout 2000 and recovers more quickly but shows marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009, and 2016. The seasonal signal is lost (see Figure \ref{fig:divts} $f$). 
Community evenness, as signified by the Pielou index, shows no seasonal signal and a very large variance within seasons and across depth (see Figure \ref{fig:divts} $h$). The inter-annual mean dynamics of evenness are remarkably stable and similar across all depth intervals, clustered around an intermediate 0.5 value. There are marked reductions in all depths towards 2000 and for all depths above 75 meters in 2006, 2009 and 2016. Throughout 2003 to 2013 evenness is almost consistently highest in the deepest interval below 75 meters, where the community seems to have been more evenly distributed.

% Figure 4
Taking the microscopy community data interpolated within the top 100 meters, we aimed to identify any apparent structure within the time series that could be related to the physical environment. To achieve this, we employed clustering analysis on both the annual aggregate of cell counts from microscopy and the individual monthly measurements. Figure \ref{fig:clustering} $a$ presents the results of clustering the annual aggregate communities using a linear complete linkage algorithm based on the binary Jaccard matrix. The community is divided into two groups, with the first cluster encompassing the years 1996-2004 and 2014-2016, while the second cluster includes the intermediate years from 2004-2013. 
Inter-annual species turnover is illustrated in Figure \ref{fig:clustering} $b$, revealing an increasing trend in turnover until 2004. Turnover subsequently decreased during the period of cluster 2, but increased again in 2014, with return to cluster 1. This pattern is also reflected in the hierarchical clustering results in Figure \ref{fig:clustering} $a$, where cluster 2 shows a lower distance in the clustering tree between the years.

To check whether the annual aggregation of community data smoothed out strong seasonal signals, we performed an NMDS cluster analysis using the monthly cell counts (see Figure \ref{fig:clustering} $c$). This NMDS clustering is overlaid with environmental data and climate indices to help explain some of the underlying trends. The first axis separates the two clusters, with cluster 2 trending towards positive anomalies in the AMO and precipitation, while cluster 1 trends towards positive anomalies in evaporation, salinity, and the MEI v.2 index. The second NMDS axis appears to follow the trends in upwelling intensity, as indicated by the opposing trend arrows for nutrient concentrations, temperature, and wind speed. Overall, NMDS clustering seems to confirm the pattern found by the annual aggregate clustering.

% Figure 5 - NEW
Figure \ref{fig:clustcomp} $a$ to $o$ shows density distributions for environmental variables, climate, and diversity indices between the two clusters. Upwelling-related variables show distinct patterns between the two clusters. For example, cluster 1 experiences more frequent high wind events, a shallower \qty{21}{\celsius} isotherm, lower temperatures, reduced evaporation, and higher salinity and nutrient concentrations, particularly for nitrate. The AMO index indicates a separation between the two clusters, with cluster 1 exhibiting more negative anomalies, while cluster 2 has a higher occurrence of positive anomalies. Although there are no significant differences in genus richness and the Shannon index between clusters, there is a slightly increased median in the Pielou index for cluster 2. The density distribution reveals some differences. For cluster 1, there is higher variance in genus richness, while cluster 2 shows greater variance and a bimodal distribution for the Shannon and Pielou indices.

% Figure 6
To identify the variables that best predict the shifts observed within the phytoplankton community, we performed a Gradient Forest analysis on the microscopy cell counts interpolated over the top 100 meters and grouped by identified genus. Preliminary model runs were used to determine if time lags of the predictor variables could better explain the data (see Appendix), resulting in the selection of the AMO with a lag of 2 months and MEI v.2 with a lag of 4 months. In Figure \ref{fig:GF} $a$, we show the predictor variables ranked by their weighted importance, with the AMO and MEI v.2 indices with time lags, as well as upwelling-related variables such as SST and nitrate in the upper 100 meters, demonstrating the highest predictive capacity. The \qty{21}{\celsius} isotherm, precipitation, and phosphate concentrations scored the lowest in predicting changes within the community. Gradient Forest predicts shifts in the community by building a random forest model for each observed genus across the time series. The predictive power of the models for each genus is shown in Figure \ref{fig:GF} $b$. The microscopy data are biased toward larger cell sizes, and therefore, most of the predicted genus cell counts are from diatoms. However, two haptophyte and three dinoflagellate genera rank relatively high.
Figure \ref{fig:GF} $c$ shows the cumulative importance of the predictor variables across the parameter range. The black line indicates cumulative importance across the entire community, and the colored dashed lines indicate importance per phytoplankton group. For the AMO index, we see a clear pattern in which negative anomalies affect the community. This contrasts with a higher importance for positive anomalies in the MEI v.2 index, particularly for haptophytes and cyanobacteria. Nitrate concentrations between 6 and 8 \unit{\micro \mole} represent the range where the community changed most, indicating these values as a threshold for community composition. See figure \ref{fig:sup:GFoutput_lags_extra} in the appendix for response curves for all variables not shown here. For wind speed, SST, and salinity, changes occurred most at parameter ranges related to high upwelling, i.e., high wind (between -4 and -6 \unit{m.s^{-1}}), low temperature (below \qty{23}{\celsius}) and high salinity (above \qty{36.8}{PSU}).

@@

\begin{figure}
\begin{center}
\noindent\includegraphics[width=480pt]{fig/DepthDivTimeSeries_updated.pdf}
\end{center}
\caption{Time series and seasonality of chlorophyll \textit{a} and diversity indices across four depth intervals. (a,c,e,g) Annual mean across time series for years with more than 5 monthly sampling. Transparent dots show monthly raw values. (b,d,f,h) Boxplots of monthly measurements across four depth intervals grouped by season. Upwelling season lasts from December to May, Rainy Season from June to November.}
\label{fig:divts}
\end{figure}

@@

This study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which is influenced by large-scale climatic changes. In the Cariaco Basin, decadal-scale climate oscillations have impacted the wind-driven upwelling regime, resulting in confounding effects on the composition and biodiversity of the phytoplankton community.
% - a few sentences on basic seasonal dynamics removed here
These changes are well reflected in the yearly mean anomalies (see Figure \ref{fig:zscore}) and the density distribution of monthly measurements (see Figure \ref{fig:clustering}). After 2004, we observed increased sea surface temperatures (SST), decreased wind speeds, reduced fluorometric chlorophyll \textit{a}, and a collapse in functional group cell counts, all corresponding to a shift in the Atlantic Multidecadal Oscillation (AMO) toward positive mean anomalies until 2014. 
% - a few sentences about shift to smaller taxa and collapse in Sardine Fisheries
